High-level cognition is supported by information-rich but compressible brain activity patterns

### 2023-12-26 paper/main.tex

\textbf{Significance Statement:} How our brains respond to ongoing experiences
depends on what we are doing and thinking about, how engaged we are, and our
past experiences, among other factors. To gain insights into how brain activity
patterns reflect ongoing cognitive processes, we examined two fundamental
aspects of brain activity under different cognitive circumstances:
informativeness and compressibility. Informativeness refers to the extent to
which brain patterns are both temporally specific and consistent across
different people. Compressibility refers to the extent to how robust the
informativeness of brain patterns is to dimensionality reduction. We found that
brain activity patterns evoked by higher-level cognitive tasks are both more
informative \textit{and} more compressible than activity evoked by lower-level
tasks. Our findings suggest that our brains flexibly reconfigure themselves to
optimize different aspects of how they function according to ongoing cognitive
demands.

@@

Large-scale networks, including the human brain, may be conceptualized as
occupying one or more positions along on a continuum. At one extreme, every
node is fully independent from every other node. At the other extreme, all
nodes behave identically. Each extreme optimizes key properties of how the
network functions. When every node is independent, the network is maximally
\textit{expressive}: if we define the network's ``state'' as the activity
pattern across its nodes, then every state is equally reachable by a network
with fully independent nodes. On the other hand, a network of identically
behaved nodes optimizes \textit{robustness}: any subset of nodes may be removed
from the network without any loss of function or expressive power, as long as
any single node remains. Presumably, most natural systems tend to occupy
positions between these extremes. We wondered: might the human brain
reconfigure itself to be more flexible or more robust according to ongoing
demands? In other words, might the brain reconfigure its connections or
behaviors under different circumstances to change its position along this
continuum?

@@

  \caption{\textbf{Information content and compressibility.} \textbf{A.
  Variance explained for two images.} Consider two example images: a photograph
  and simple drawing. The images have the same resolutions, but their content
  is very different. The photograph is more information rich: when we compare
  the original photograph and drawing (leftmost column), we can see that the
  photograph captures much more detail. We can also apply principal components
  analysis to the images, treating the rows of the images as ``observations.''
  Across columns, we identified the numbers of components required to explain
  100\%, 95\%, 75\%, or 50\% of the cumulative variance in each image (the
  100\% columns denote the original images). The numbers of components are
  indicated in parentheses, and the resulting ``compressed'' images are
  displayed. This analysis reveals that the drawing is more compressible: just
  16 components can explain 95\% of the variance in teh drawing, whereas 50
  components are required to explain 95\% of the variance in the photograph.
  \textbf{B. Representing two images with different numbers of components.}
  Using the same principal component decompositions as in Panel A, we computed
  the cumulative proportion of variance explained with 380 (original images),
  20, 4, or 2 components. This analysis provides another way of characterizing
  the compressibility of each image by showing that the same number of
  components can explain more variance in the drawing than in the photograph.
  \textbf{C. Template data from four synthetic datasets.} We constructed four
  synthetic datasets, each comprising 25 features (columns) observed across 100
  samples (rows) from each of 10 simulated ``participants.'' The datasets were
  constructed to contain different levels of informativeness and
  compressibility (see \textit{Synthetic data}). \textbf{D. Decoding accuracy
  by number of components.} For each synthetic dataset, we trained
  across-participant classifiers to decode timepoint labels.  Each panel displays the
  decoding accuracy as a function of the number of components used to represent
  the data. Error ribbons denote bootstrap-estimated 95\% confidence intervals.}
\label{fig:information-compression} 
\end{figure}

@@

To evaluate the relation between informativeness and compressibility for brain
activity from each experimental condition, we trained a series of
across-participant temporal decoders on compressed representations of the data.
Figure~\ref{fig:inflection}A displays the decoding accuracy as a function of
the number of principal components used to represent the data (also see
Fig.~\varExplained). Several patterns were revealed by the analysis. First, in
general (i.e., across experimental conditions), decoding accuracy tends to
improve as the number of components are increased. However, decoding accuracy
peaked at higher levels for experimental conditions that exposed participants
to cognitively richer stimuli (Fig.~\ref{fig:inflection}D). The peak decoding
accuracy was highest for the ``intact'' condition (versus paragraph: $t(99) =
35.205, p < 0.001)$; versus word: $t(99) = 43.172, p < 0.001)$; versus rest:
$t(99) = 81.361, p < 0.001)$), next highest for the ``paragraph'' condition
(versus word: $t(99) = 6.243, p < 0.001)$; versus rest: $t(99) = 50.748, p <
0.001)$), and next highest for the ``word'' condition (versus rest: $t(99) =
48.791, p < 0.001)$). This ordering implies that cognitively richer conditions
evoke more stable brain activity patterns across people.

@@

These results make some intuitive sense. As the contextual information
available to participants increases (i.e., over time in the cognitively rich
intact and paragraph conditions), it makes sense that this might
constrain neural responses to a greater extent. While this pattern may not
necessarily hold for \textit{every} possible story or stimulus, we suspect that
it is generally the case that our knowledge about what is happening in a story
tends to increase as we experience more about it. In turn, this could lead to
greater consistency in different people's interpretations of and neural
responses to the stimulus. Similarly, as participants are left to ``mind
wander,'' or as they experience mental fatigue (i.e., over time in the less
cognitively rich word and rest conditions), we suggest that this might
lead to greater variability in neural responses across people, resulting in
lower decoding accuracy. Again, it is not necessarily the case that every
possible ``unengaging'' stimulus will lead to greater neural variability as
time progresses, but we suspect this phenomenon is likely to hold for a variety
of such stimuli. These findings replicate at least to a limited extent (e.g.,
across both cognitively rich conditions and across both cognitively
impoverished conditions, and for the different groups of participants in each
of those conditions). However, determining whether these patterns generalize to
other stimuli would require additional study (with new stimuli).

@@

If the informativeness and compressibility of brain activity patterns vary over
time, do these properties might also vary across brain networks? We used a
network parcellation identified by \cite{YeoEtal11} to segment the brain into
seven distinct networks. The networks can be sorted (roughly) in order from
lower-level to higher-level cortex as follows (Figs.~\ref{fig:networks}A--C):
visual, somatomotor, dorsal attention, ventral attention, limbic,
frontoparietal, and default mode. Next, we computed decoding curves separately
for the activity patterns within each network and identified each network's
inflection point, for each experimental condition. Moving from low-order
networks to higher-order networks, we found that decoding accuracy tended to
increase in the higher-level experimental conditions and decrease (slightly) in
the lower-level experimental conditions (Fig.~\ref{fig:networks}D, E;
Spearman's rank correlation between decoding accuracy and network order:
intact: $\rho = 0.362, p < 0.001$; paragraph: $\rho = 0.441, p < 0.001$; word:
$\rho = -0.102, p = 0.007$; rest: $\rho = -0.354, p < 0.001$). This suggests
that higher-order networks may carry more content-relevant or stimulus-driven
``information.'' We found no clear trends in the proportions of components
required to achieve 5\% decoding accuracy across networks or conditions
(Fig.~\ref{fig:networks}F). We note that the limbic network we considered here
often overlaps with low (imaging) signal regions, and therefore it may be
difficult to draw strong conclusions about this network's informativeness or
compressibility. We also considered the possibility that the correlations with
network order might be influenced by the numbers of nodes in each network. We
designed a permutation-based procedure to address this possibility, whereby we
repeated the above analyses using shuffled network labels (see \textit{Network
permutation tests}). The correlations between decoding accuracy and network
order were reliably more positive than the shuffled correlations for the intact
($t(1998) = 276.431, p < 0.001$) and paragraph ($t(1998) = 330.334, p < 0.001$)
conditions, and reliably more negative for the word ($t(1998) = -16.386, p <
0.001$) and rest ($t(1998) = -318.631, p < 0.001$) conditions. These results
suggest that the correlations between decoding accuracy and network order were
not driven solely by the numbers of nodes in each network.

@@

Whereas the above analyses examinine different networks in isolation, how does
full-brain (i.e., potentially multi-network) activity patterns reflected by
different principal components vary across different experimental conditions?
As shown in Figure~\ref{fig:neurosynth-pca}, we used
Neurosynth~\citep{RubiEtal17} to identify, for each component, the associations
with each of 80 themes (see \textit{Reverse inference}). In general, the first
principal components across all of the experimental conditions tended to weight
most heavily on themes related to cognitive control, memory, language
processing, attention, and perception. Other components appeared to vary more
across conditions.

To gain further insights into which brain functions might be most closely
associated with the top-weighted components from each experimental condition,
we manually grouped each Neurosynth-derived topic into a smaller set of core
cognitive functions. Separately for each component, we computed the average
weightings across all topics that were tagged as being associated with each of
these cognitive functions (Figs.~\ref{fig:neurosynth-summary}A,~\neurosynthFull
A). To help visualize these associations, we used the patterns of associations
for each component to construct graphs whose nodes were experimental conditions
and cognitive functions (Figs.~\ref{fig:neurosynth-summary}B,~\neurosynthFull
B). We also computed correlations between the sets of per-topic weightings from
each of the top-weighted components from each experimental condition
(Fig.~\ref{fig:neurosynth-summary}C, D) and between the brain maps for each
condition's components (Fig.~\neurosynthFull C, D). Taken together, we found
that each component appeared to weight on a fundamental set of cognitive
functions that varied by experimental condition. For example, the top principal
components from every condition weighted similarly (across conditions) on the
full set of Neurosynth topics (Fig.~\ref{fig:neurosynth-pca}A) and cognitive
functions (Figs.~\ref{fig:neurosynth-summary}A, B and~\neurosynthFull~A, B),
suggesting that these components might reflect a set of functions or activity
patterns that are common across all conditions. The second components'
weightings were similar across the intact, paragraph, and rest conditions
(highest-weighted functions: cognitive control, memory, social cognition, and
resting state), but different for the word condition (highest-weighted
functions: sensory perception and cognitive control). The fourth components'
weighting grouped the paragraph and word conditions (highest-weighted
functions: memory, language processing, and cognitive control) and the intact
and rest conditions (highest-weighted functions: emotion, social cognition). We
also used ChatGPT~\citep{ChatGPT} to sort the list of manually tagged cognitive
functions from lowest-level to highest-level (Tab.~\topicTags,
Fig.~\ref{fig:neurosynth-summary}E; also see \textit{Ranking cognitive
processes}). We found that higher-level functions tended to be weighted on more
heavily by top components from the intact and paragraph conditions than
lower-level functions (intact vs. word: $t(198) = 11.059, p < 0.001$; intact
vs. rest: $t(198) = 3.699, p < 0.001$; paragraph vs. word: $t(198) = 13.504, p
< 0.001$; paragraph vs. rest: $t(198) = 4.812, p < 0.001$; also see
\textit{Ranking cognitive processes}). The top components from the word
condition showed the opposite tendency, whereby \textit{lower}-level functions
tended to be weighted on more heavily than higher-level functions (word vs.
rest: $t(198) = -7.315, p < 0.001$). The weighting trends for the intact and
paragraph conditions were not reliably different ($t(198) = -0.479, p =
0.633$). The components from the rest condition showed almost no differences in
the weights associated with high-level versus low-level functions (rest vs. 0:
$t(99) = 1.836, p = 0.081$). These findings suggest that when participants were
engaged more strongly (in the more engaging intact and paragraph conditions),
their dominant neural patterns reflected higher-level cognitive functions. In
contrast, when participants were engaged less strongly (in the less engaging
word and rest conditions), their dominant neural patterns reflected lower-level
cognitive functions. Although they were highly statistically reliable, it is
also important to note that these latter effects are also relatively small
(e.g., the slopes for \textit{all} of the experimental conditions are
numerically close to zero; Fig.~\ref{fig:neurosynth-summary}E). We suggest that
this phenomenon may merit further investigation in future work.

@@

\caption{\textbf{Summary of functions associated with top-weighted components
by condition.} \textbf{A. Top-weighted topics by condition.} Here we display
per-condition (rows, indicated by colored dots) topic correlations, averaged
across topics that pertain to each of several broad cognitive functions
(columns within each sub-panel, indicated by colored dots). Each sub-panel
reflects correlations for the components indicated in the panel titles. A
legend for the condition and several core cognitive function classifications is
displayed in the lower right of the figure. Table~\topics~provides a list of
each topic's top-weighted terms, along with each topic's manually labeled
cognitive classification. A full list of the topics most highly associated with
each component may be found in Figure~\topTerms. \textbf{B. Associations
between per-condition components and cognitive functions.} The network plots
denote positive average correlations between the component images for each
condition (gray-outlined dots on the left sides of each network; colors denote
conditions) and topic-specific brain maps associated with each indicated
cognitive function (black-outlined dots on the right sides of each network;
colors denote cognitive functions). The line thicknesses are proportional to
the correlation values (correlation coefficients are noted in the heat maps in
Panel A). \textbf{C. Correlations between each principal component, by
condition.} The heat maps display the correlations between the brain maps
(Fig.~\componentBrains) for each principal component (sub-panel), across each
pair of conditions (rows and columns of each sub-panel's matrix, indicated by
colored dots). \textbf{D. Associations between per-condition components, by
component.} Each sub-panel's network plot summarizes the pattern of
correlations between the $n$\textsuperscript{th} top-weighted principal
components (sub-panel) for each experimental condition (gray-outlined dots).
The line thicknesses are proportional to the correlation values (correlation
coefficients are noted in the heat maps in Panel C). \textbf{E. Change in
weights by condition.} The bar heights reflect the slopes of regression lines,
fit separately to the top five components from each condition, between the
ChatGPT-derived ``rank'' of each cognitive classification and the correlations
between the component and topic maps associated with cognitive processes at the
given rank (see \textit{Ranking cognitive processes}) Error bars denote
bootstrap-estimated 95\% confidence intervals. Also see
Fig.~\neurosynthFull~for additional information.}

@@

We examined fMRI data collected as participants listened to an auditory
recording of a story, scrambled recordings of the story, or underwent a resting
state scan. We found that cognitively richer stimuli evoked more reliable
(i.e., consistent across people) and information rich brain activity patterns.
The brain patterns evoked by cognitively richer stimuli were also more
compressible, in that each individual component provided more ``signal'' to
temporal decoders relative to components of data from less cognitively rich
conditions (Fig.~\ref{fig:inflection}). Over time (e.g., as the experiment
progressed), these phenomena were strengthened. Specifically, across story
segments, data from more cognitively rich conditions became more informative
and compressible, and data from less cognitively rich conditions became
\textit{less} informative and compressible
(Fig.~\ref{fig:inflection-quarters}).  We also repeated these analyses
separately for different brain networks. We found that networks traditionally
associated with higher-level cognitive functions tended to provide more
informative brain patterns than networks traditionally associated with lower
level cognitive functions (Fig.~\ref{fig:networks}). Finally, we examined the
most dominant components of the brain activity patterns from each experimental
condition. We used a reverse inference approach~\citep{RubiEtal17} to identify
the terms in the neuroimaging literature most commonly associated with the
corresponding maps. As summarized in Figure~\ref{fig:neurosynth-summary}E, we
found that the intact and paragraph conditions tended to weight on higher-level
cognitive processes more than lower-level cognitive processes, whereas the word
condition weighted on lower-level processes more than higher-level processes
and the rest condition showed no difference in high-level versus low-level
weighting. Taken together, our findings indicate that the informativeness and
compressibility of our brain activity patterns are task-dependent, and these
properties change systematically with factors like cognitive richness and depth
of processing.

Our explorations of informativeness and compressibility are related to a much
broader literature on the correlational and causal structure of brain activity
patterns and networks~\citep{PretEtal17, OwenEtal21, RogeEtal07, RubiSpor10,
SizeEtal18, SmitEtal13b, SmitEtal13c, SrinEtal07, TomaVolk11, YeoEtal11,
AdacEtal12, BassSpor17, BullSpor09, SporHone06, SporBetz16, SporZwi04,
DhamEtal08, KorzEtal08, BrovEtal04, LynnBass21}. Correlations or causal
associations between different brain regions simultaneously imply that
full-brain activity patterns will be compressible and also that those activity
patterns will contain redundancies. For example, the extent to which activity
patterns at one brain area can be inferred or predicted from activity patterns
at other areas~\citep[e.g.,][]{OwenEtal20, ScanEtal21}, reflects overlap in the
information available in or represented by those brain areas. If brain patterns
in one area are recoverable using brain patterns in another area, then a
``signal'' used to convey the activity patterns could be compressed by removing
the recoverable activity. Predictable (and therefore redundant) brain activity
patterns are also more robust to signal corruption. For example, even if the
activity patterns at one region are unreadable or unreliable at a given moment,
that unreliability could be compensated for by other regions' activity patterns
that were predictive of the unreliable region. Whereas compressible brain
patterns a robust to spatial signal corruption, high versus low informativeness
reflects a similar (though dissociable; e.g.,
Fig.~\ref{fig:information-compression}) tradeoff between expressiveness and
robustness of \textit{temporal} patterns. Highly informative brain patterns (by
our measure; i.e., patterns that yield greater temporal decoding accuracy) are
expressive about ongoing experiences or cognitive states, since each moment's
pattern is reliably distinguishable from other moments' patterns. However, when
each moment's pattern is unique, brain activity becomes less robust to temporal
signal corruption. Our finding that brain activity patterns becomes more
informative (i.e., less robust to temporal signal corruption) and compressible
(i.e., more robust to spatial signal corruption) when cognitive engagement is
higher suggests that our brain may optimize its activity patterns to prioritize
either temporal or spatial robustness, according to task demands.

@@

We found that different brain networks varied in how informative and
compressible their activity patterns were across experimental conditions (e.g.,
Fig.~\ref{fig:networks}). This might
follow from evolutionary optimizations that reflect the relevant constraints or
demands placed on those networks. One possibility is that cortex is organized
in a hierarchy of networks ``concerned with'' or selective to different levels
of processing or function. To the extent that different levels of processing
(e.g., low-level sensory processing versus ``deeper'' higher-level processing)
reflect different stimulus timescales~\citep[e.g.,][]{Mann20}, the network
differences we observed might also relate to the timescales at which each
network is maximally sensitive~\citep{RegeEtal18, BaldEtal17,LernEtal11,
HassEtal08}.

>>>>

\textbf{Significance Statement:} How our brains respond to ongoing
experiences depends on what we are doing and thinking about, among
other factors. To study how brain activity reflects ongoing cognition,
we examined two fundamental aspects of brain activity under different
cognitive circumstances: informativeness and
compressibility. Informativeness refers to the extent to which brain
patterns are both temporally specific and consistent across different
people. Compressibility refers to how robust the informativeness of
brain patterns is to dimensionality reduction. Brain activity evoked
by higher-level cognitive tasks are both more informative \textit{and} more
compressible than activity evoked by lower-level tasks. Our findings
suggest that our brains flexibly reconfigure themselves to optimize
different aspects of how they function according to ongoing cognitive
demands.

@@

Large-scale networks, including the human brain, may be conceptualized as
occupying one or more positions along on a continuum. At one extreme, every
node is fully independent from every other node. At the other extreme, all
nodes behave identically. Each extreme optimizes key properties of how the
network functions. When every node is independent, the network is maximally
\textit{expressive}: if we define the network's ``state'' as the activity
pattern across its nodes, then every state is equally reachable by a network
with fully independent nodes. On the other hand, a network of identically
behaved nodes optimizes \textit{robustness}: any subset of nodes may be removed
from the network without any loss of function or expressive power, as long as
any single node remains. In addition to considering flexibility across space
(nodes), these properties may also vary, largely independently, across time. A
network is maximally expressive when its nodes' activity patterns vary in
meaningful ways from moment to moment, whereas it is maximally robust to signal
corruption when its activity is constant over time. Presumably, most natural
systems tend to occupy positions between these temporal and spatial extremes. Under
different circumstances, it may even prove benefitial for systems to make
different tradeoffs between expressiveness and robustness along the temporal
and spatial dimensions.  We wondered: might the human brain reconfigure itself to
be more flexible or more robust according to ongoing demands? In other words,
might the brain reconfigure its connections or behaviors under different
circumstances to change its position along these continuums?

@@

  \caption{\textbf{Information content and compressibility.} \textbf{A.
  Variance explained for two images.} Consider two example images: a photograph
  and simple drawing. The images have the same resolutions, but their content
  is very different. The photograph is more information rich: when we compare
  the original photograph and drawing (leftmost column), we can see that the
  photograph captures much more detail. We can also apply principal components
  analysis to the images, treating the rows of the images as ``observations.''
  Across columns, we identified the numbers of components required to explain
  100\%, 95\%, 75\%, or 50\% of the cumulative variance in each image (the
  100\% columns denote the original images). The numbers of components are
  indicated in parentheses, and the resulting ``compressed'' images are
  displayed. This analysis reveals that the drawing is more compressible: just
  16 components can explain 95\% of the variance in the drawing, whereas 50
  components are required to explain 95\% of the variance in the photograph.
  \textbf{B. Representing two images with different numbers of components.}
  Using the same principal component decompositions as in Panel A, we computed
  the cumulative proportion of variance explained with 380 (original images),
  20, 4, or 2 components. This analysis provides another way of characterizing
  the compressibility of each image by showing that the same number of
  components can explain more variance in the drawing than in the photograph.
  \textbf{C. Template data from four synthetic datasets.} We constructed four
  synthetic datasets, each comprising 25 features (columns) observed across 100
  samples (rows) from each of 10 simulated ``participants.'' The datasets were
  constructed to contain different levels of informativeness and
  compressibility (see \textit{Synthetic data}). \textbf{D. Decoding accuracy
  by number of components.} For each synthetic dataset, we trained
  across-participant classifiers to decode timepoint labels.  Each panel displays the
  decoding accuracy as a function of the number of components used to represent
  the data. Error ribbons denote bootstrap-estimated 95\% confidence intervals.}
\label{fig:information-compression} 
\end{figure}

@@

To evaluate the relation between informativeness and compressibility for brain
activity from each experimental condition, we trained a series of
across-participant temporal decoders on compressed representations of the data.
Figure~\ref{fig:inflection}A displays the decoding accuracy as a function of
the number of principal components used to represent the data (also see
Fig.~\varExplained). Several patterns were revealed by the analysis. First, in
general (i.e., across experimental conditions), decoding accuracy tends to
improve as the number of components are increased. However, decoding accuracy
peaked at higher levels for experimental conditions that exposed participants
to cognitively richer stimuli (Fig.~\ref{fig:inflection}D). The peak decoding
accuracy was highest for the ``intact'' condition (versus paragraph: $t(99) =
35.205, p < 0.001$; versus word: $t(99) = 43.172, p < 0.001$; versus rest:
$t(99) = 81.361, p < 0.001$), next highest for the ``paragraph'' condition
(versus word: $t(99) = 6.243, p < 0.001$; versus rest: $t(99) = 50.748, p <
0.001$), and next highest for the ``word'' condition (versus rest: $t(99) =
48.791, p < 0.001$). This ordering implies that cognitively richer conditions
evoke more stable brain activity patterns across people.

@@

These results make some intuitive sense. As the contextual information
available to participants increases (i.e., over time in the cognitively rich
intact and paragraph conditions), it makes sense that this might
constrain neural responses to a greater extent. While this pattern may not
necessarily hold for \textit{every} possible story or stimulus, we suspect that
it is generally the case that our knowledge about what is happening in a story
tends to increase as we experience more of it. In turn, this could lead to
greater consistency in different people's interpretations of and neural
responses to the stimulus. Similarly, as participants are left to ``mind
wander,'' or as they experience mental fatigue (i.e., over time in the less
cognitively rich word and rest conditions), we suggest that this might
lead to greater variability in neural responses across people, resulting in
lower decoding accuracy. Again, it is not necessarily the case that every
possible ``unengaging'' stimulus will lead to greater neural variability as
time progresses, but we suspect this phenomenon is likely to hold for a variety
of such stimuli. These findings replicate at least to a limited extent (e.g.,
across both cognitively rich conditions and across both cognitively
impoverished conditions, and for the different groups of participants in each
of those conditions). However, determining whether these patterns generalize to
other stimuli would require additional study (with new stimuli).

@@

If the informativeness and compressibility of brain activity patterns vary over
time, might these properties also vary across brain networks? We used a
network parcellation identified by \cite{YeoEtal11} to segment the brain into
seven distinct networks. The networks can be sorted (roughly) in order from
lower-level to higher-level cortex as follows (Figs.~\ref{fig:networks}A--C):
visual, somatomotor, dorsal attention, ventral attention, limbic,
frontoparietal, and default mode. Next, we computed decoding curves separately
for the activity patterns within each network and identified each network's
inflection point, for each experimental condition. Moving from low-order
networks to higher-order networks, we found that decoding accuracy tended to
increase in the higher-level experimental conditions and decrease (slightly) in
the lower-level experimental conditions (Fig.~\ref{fig:networks}D, E;
Spearman's rank correlation between decoding accuracy and network order:
intact: $\rho = 0.362, p < 0.001$; paragraph: $\rho = 0.441, p < 0.001$; word:
$\rho = -0.102, p = 0.007$; rest: $\rho = -0.354, p < 0.001$). This suggests
that higher-order networks may carry more content-relevant or stimulus-driven
``information.'' We found no clear trends in the proportions of components
required to achieve 5\% decoding accuracy across networks or conditions
(Fig.~\ref{fig:networks}F). We note that the limbic network we considered here
often overlaps with low (imaging) signal regions, and therefore it may be
difficult to draw strong conclusions about this network's informativeness or
compressibility. We also considered the possibility that the correlations with
network order might be influenced by the numbers of nodes in each network. We
designed a permutation-based procedure to address this possibility, whereby we
repeated the above analyses using shuffled network labels (see \textit{Network
permutation tests}). The correlations between decoding accuracy and network
order were reliably more positive than the shuffled correlations for the intact
($t(1998) = 276.431, p < 0.001$) and paragraph ($t(1998) = 330.334, p < 0.001$)
conditions, and reliably more negative for the word ($t(1998) = -16.386, p <
0.001$) and rest ($t(1998) = -318.631, p < 0.001$) conditions. These results
suggest that the correlations between decoding accuracy and network order were
not driven solely by the numbers of nodes in each network.

@@

Whereas the above analyses examined different networks in isolation, how does
full-brain (i.e., potentially multi-network) activity patterns reflected by
different principal components vary across different experimental conditions?
As shown in Figure~\ref{fig:neurosynth-pca}, we used
Neurosynth~\citep{RubiEtal17} to identify, for each component, the associations
with each of 80 themes (see \textit{Reverse inference}). In general, the first
principal components across all of the experimental conditions tended to weigh
most heavily on themes related to cognitive control, memory, language
processing, attention, and perception. Other components appeared to vary more
across conditions.

To gain further insights into which brain functions might be most closely
associated with the top-weighted components from each experimental condition,
we manually grouped each Neurosynth-derived topic into a smaller set of core
cognitive functions. Separately for each component, we computed the average
weightings across all topics that were tagged as being associated with each of
these cognitive functions (Figs.~\ref{fig:neurosynth-summary}A,~\neurosynthFull
A). To help visualize these associations, we used the patterns of associations
for each component to construct graphs whose nodes were experimental conditions
and cognitive functions (Figs.~\ref{fig:neurosynth-summary}B,~\neurosynthFull
B). We also computed correlations between the sets of per-topic weightings from
each of the top-weighted components from each experimental condition
(Fig.~\ref{fig:neurosynth-summary}C, D) and between the brain maps for each
condition's components (Fig.~\neurosynthFull C, D). Taken together, we found
that each component appeared to weigh on a fundamental set of cognitive
functions that varied by experimental condition. For example, the top principal
components from every condition weighed similarly (across conditions) on the
full set of Neurosynth topics (Fig.~\ref{fig:neurosynth-pca}A) and cognitive
functions (Figs.~\ref{fig:neurosynth-summary}A, B and~\neurosynthFull~A, B),
suggesting that these components might reflect a set of functions or activity
patterns that are common across all conditions. The second components'
weightings were similar across the intact, paragraph, and rest conditions
(highest-weighted functions: cognitive control, memory, social cognition, and
resting state), but different for the word condition (highest-weighted
functions: sensory perception and cognitive control). The fourth components'
weighting grouped the paragraph and word conditions (highest-weighted
functions: memory, language processing, and cognitive control) and the intact
and rest conditions (highest-weighted functions: emotion, social cognition). We
also used ChatGPT~\citep{ChatGPT} to sort the list of manually tagged cognitive
functions from lowest-level to highest-level (Tab.~\topicTags,
Fig.~\ref{fig:neurosynth-summary}E; also see \textit{Ranking cognitive
processes}). We found that higher-level functions tended to be weighted more
heavily by top components from the intact and paragraph conditions than
lower-level functions (intact vs. word: $t(198) = 11.059, p < 0.001$; intact
vs. rest: $t(198) = 3.699, p < 0.001$; paragraph vs. word: $t(198) = 13.504, p
< 0.001$; paragraph vs. rest: $t(198) = 4.812, p < 0.001$; also see
\textit{Ranking cognitive processes}). The top components from the word
condition showed the opposite tendency, whereby \textit{lower}-level functions
tended to be weighted more heavily than higher-level functions (word vs. rest:
$t(198) = -7.315, p < 0.001$). The weighting trends for the intact and
paragraph conditions were not reliably different ($t(198) = -0.479, p =
0.633$). The components from the rest condition showed only a small trending
difference in the weights associated with high-level versus low-level functions
(rest vs. 0: $t(99) = 1.836, p = 0.081$). These findings suggest that when
participants were engaged more strongly (in the more engaging intact and
paragraph conditions), their dominant neural patterns reflected higher-level
cognitive functions. In contrast, when participants were engaged less strongly
(in the less engaging word and rest conditions), their dominant neural patterns
reflected lower-level cognitive functions. Although they were highly
statistically reliable, it is also important to note that these latter effects
are also relatively small (e.g., the slopes for \textit{all} of the
experimental conditions are numerically close to zero;
Fig.~\ref{fig:neurosynth-summary}E). We suggest that this phenomenon may merit
further investigation in future work.

@@

\caption{\textbf{Summary of functions associated with top-weighted components
by condition.} \textbf{A. Top-weighted topics by condition.} Here we display
per-condition (rows, indicated by colored dots) topic correlations, averaged
across topics that pertain to each of several broad cognitive functions
(columns within each sub-panel, indicated by colored dots). Each sub-panel
reflects correlations for the components indicated in the panel titles.
Table~\topics~provides a list of each topic's top-weighted terms, along with
each topic's manually labeled cognitive classification. A full list of the
topics most highly associated with each component may be found in
Figure~\topTerms. \textbf{B. Associations between per-condition components and
cognitive functions.} The network plots denote positive average correlations
between the component images for each condition (gray-outlined dots on the left
sides of each network; colors denote conditions) and topic-specific brain maps
associated with each indicated cognitive function (black-outlined dots on the
right sides of each network; colors denote cognitive functions). The line
thicknesses are proportional to the correlation values (correlation
coefficients are noted in the heat maps in Panel A). \textbf{C. Correlations
between each principal component, by condition.} The heat maps display the
correlations between the brain maps (Fig.~\componentBrains) for each principal
component (sub-panel), across each pair of conditions (rows and columns of each
sub-panel's matrix, indicated by colored dots). \textbf{D. Associations between
per-condition components, by component.} Each sub-panel's network plot
summarizes the pattern of correlations between the $n$\textsuperscript{th}
top-weighted principal components (sub-panel) for each experimental condition
(gray-outlined dots). The line thicknesses are proportional to the correlation
values (correlation coefficients are noted in the heat maps in Panel C).
\textbf{E. Change in weights by condition.} The bar heights reflect the slopes
of regression lines, fit separately to the top five components from each
condition, between the ChatGPT-derived ``rank'' of each cognitive
classification and the correlations between the component and topic maps
associated with cognitive processes at the given rank (see \textit{Ranking
cognitive processes}) Error bars denote bootstrap-estimated 95\% confidence
intervals. Also see Fig.~\neurosynthFull~for additional information.}

@@

We examined fMRI data collected as participants listened to an auditory
recording of a story, scrambled recordings of the story, or underwent a resting
state scan. We found that cognitively richer stimuli evoked more reliable
(i.e., consistent across people) and information rich brain activity patterns.
The brain patterns evoked by cognitively richer stimuli were also more
compressible, in that each individual component provided more ``signal'' to
temporal decoders relative to components of data from less cognitively rich
conditions (Fig.~\ref{fig:inflection}). Over time (e.g., as the experiment
progressed), these phenomena were strengthened. Specifically, across story
segments, data from more cognitively rich conditions became more informative
and compressible, and data from less cognitively rich conditions became
\textit{less} informative and compressible
(Fig.~\ref{fig:inflection-quarters}).  We also repeated these analyses
separately for different brain networks. We found that networks traditionally
associated with higher-level cognitive functions tended to provide more
informative brain patterns than networks traditionally associated with lower
level cognitive functions (Fig.~\ref{fig:networks}). Finally, we examined the
most dominant components of the brain activity patterns from each experimental
condition. We used a reverse inference approach~\citep{RubiEtal17} to identify
the terms in the neuroimaging literature most commonly associated with the
corresponding maps. As summarized in Figure~\ref{fig:neurosynth-summary}E, we
found that the intact and paragraph conditions tended to weight higher-level
cognitive processes more than lower-level cognitive processes, whereas the word
condition weighted lower-level processes more than higher-level processes
and the rest condition showed no reliable difference in high-level versus low-level
weightings. Taken together, our findings indicate that the informativeness and
compressibility of our brain activity patterns are task-dependent, and these
properties change systematically with factors like cognitive richness and depth
of processing.

Our explorations of informativeness and compressibility are related to a much
broader literature on the correlational and causal structure of brain activity
patterns and networks~\citep{PretEtal17, OwenEtal21, RogeEtal07, RubiSpor10,
SizeEtal18, SmitEtal13b, SmitEtal13c, SrinEtal07, TomaVolk11, YeoEtal11,
AdacEtal12, BassSpor17, BullSpor09, SporHone06, SporBetz16, SporZwi04,
DhamEtal08, KorzEtal08, BrovEtal04, LynnBass21}. Correlations or causal
associations between different brain regions simultaneously imply that
full-brain activity patterns will be compressible and also that those activity
patterns will contain redundancies. For example, the extent to which activity
patterns at one brain area can be inferred or predicted from activity patterns
at other areas~\citep[e.g.,][]{OwenEtal20, ScanEtal21}, reflects overlap in the
information available in or represented by those brain areas. If brain patterns
in one area are recoverable using brain patterns in another area, then a
``signal'' used to convey the activity patterns could be compressed by removing
the recoverable activity. Predictable (and therefore redundant) brain activity
patterns are also more robust to signal corruption. For example, even if the
activity patterns at one region are unreadable or unreliable at a given moment,
that unreliability could be compensated for by other regions' activity patterns
that were predictive of the unreliable region. Whereas compressible brain
patterns are robust to spatial signal corruption, high versus low
informativeness reflects a similar (though dissociable; e.g.,
Fig.~\ref{fig:information-compression}) tradeoff between expressiveness and
robustness of \textit{temporal} patterns. Highly informative brain patterns (by
our measure; i.e., patterns that yield greater temporal decoding accuracy) are
expressive about ongoing experiences or cognitive states, since each moment's
pattern is reliably distinguishable from other moments' patterns. However, when
each moment's pattern is unique, brain activity becomes less robust to temporal
signal corruption. Our finding that brain activity patterns becomes more
informative (i.e., less robust to temporal signal corruption) and compressible
(i.e., more robust to spatial signal corruption) when cognitive engagement is
higher suggests that our brain may optimize its activity patterns to prioritize
either temporal or spatial robustness, according to task demands.

@@

We found that different brain networks varied in how informative and
compressible their activity patterns were across experimental conditions (e.g.,
Fig.~\ref{fig:networks}). This might
follow from evolutionary optimizations that reflect the relevant constraints or
demands placed on those networks. One possibility is that cortex is organized
in a hierarchy of networks ``concerned with'' or selective to different levels
of processing or function. To the extent that different levels of processing
(e.g., low-level sensory processing versus ``deeper'' higher-level processing)
reflect different stimulus timescales~\citep[e.g.,][]{Mann23b}, the network
differences we observed might also relate to the timescales at which each
network is maximally sensitive~\citep{RegeEtal18, BaldEtal17,LernEtal11,
HassEtal08}.
